One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-07 paper/discussion.tex
minor edits
major updates

\subsubsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}.

Considering the spiral as a external potential (c.f. an embedded
satellite), we expect it to  exert a torque on the disc at Lindblad
and co-rotation resonances.

@@

\subsection{Speculations}
%speculations - long term balance. forced temp gradient drives
%instability (disc gains ang mom, spiral loses ang mom). but if spiral
%shocks and dissiple, it dumps negative angular momentum (disc gains
%negative ang mom). balance between gain from forced temp gradient and
%loss due to dissiplatoin

>>>>

\subsubsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. To estimate this effect, we port basic results from
disc-planet theory \citep[see, e.g.][and references
therein]{papaloizou07}.

Let us regard the one-arm spiral disturbance located in the inner disc as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We expect
it to exert a torque on the exterior disc at Lindblad  
and co-rotation resonances. At the outer Lindblad resonance (OLR), this
torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
\left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
\end{align}   
where a Keperian disc has been assumed and subscript $L$ denotes
evaluation at $R=R_L$. (The inner Lindblad resonance does not exist
for the pattern speeds observed in our simulations.)

To order-of-magnitude, we reasonably take
\begin{align}
  \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \bar{R}|}, 
\end{align}
for the potential associated with a disturbance located at
$R=\bar{R}$, where $M_\mathrm{ring}$ is the mass in $R\in[R_1,R_2]$.
If we also associate its angular momentum with 
$J_\mathrm{ext}  = M_\mathrm{ring}\bar{R}^2\Omega_p$, we can obtain a
rate $\gamma_L$ by evaluating $\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
    Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)^{-1/2}\left(\frac{R_c}{\bar{R}}\right)^{5/2}\notag\\
  &\times \left\{ \frac{R_L}{R_L - \bar{R}} - 2\left[1 -
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
    - \bar{R}\right)^2}. 
\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\bar{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 0.012\Omega_p. 
\end{align}
This represents an upper limit on the effect of the OLR on the spiral
disturbance, and is much smaller than the growth rate measured in the
simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
the OLR has negligible effect on the one-arm spiral.

For the co-rotation torque, we use the result
\begin{align}
  \Gamma_C = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{C}
\end{align}
for a Keplerian disc, where subscript $C$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
  \Gamma_C = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
    \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
      - \bar{R}\right)^2}.   
\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \frac{\gamma_C}{\Omega_p} \sim 0.010\Omega_p,
\end{align}
which is again small compared to the measured growth rates.

%expressions for linear response, but have massive "planet" -> but CR
%and LR far from where spiral is

@@

%\subsection{Speculations}
%speculations - long term balance. forced temp gradient drives
%instability (disc gains ang mom, spiral loses ang mom). but if spiral
%shocks and dissiple, it dumps negative angular momentum (disc gains
%negative ang mom). balance between gain from forced temp gradient and
%loss due to dissiplatoin

### 2014-11-07 paper/model.tex
major updates

\subsubsection{Three-dimensionality}
In 3D simulations we measure the importance of vertical motion with
$\Theta$, where 
\begin{align}
  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
\end{align}
and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
\begin{align}
  \avg{v_z^2} \equiv\frac{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho v_z^2 dV}{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho dV
  },
\end{align}
and similarly for the horizontal velocities. The radial range of
integration is taken over $r\in[R_1,R_2]$ since this is where
we find the perturbations to be confined.

\section{Linear density waves}\label{wkb}
In this paper we take a fully numerical approach using non-linear
hydrodynamic simulations. However, it will be useful to employ
existing linear density wave theory to interpret the simulations.
We list here the main definitions and results from 2D linear theory,
taken from \cite{lin11} and \cite{shu91}.

@@

However, as noted in \cite{lin11}, if there is an imposed
temperature gradient, as in the disc models we consider,
then $\tbg\neq 0$ in general, which corresponds to a torque
exerted by the background disc on the perturbation. We 
outline the derivation of $T_\mathrm{BG}$ in Appendix 
\ref{tbg_deriv}.

>>>>

\subsubsection{Three-dimensionality}
In 3D simulations we measure the importance of vertical motion with
$\Theta$, where 
\begin{align}\label{theta}
  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
\end{align}
and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
\begin{align}
  \avg{v_z^2} \equiv\frac{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho v_z^2 dV}{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho dV
  },
\end{align}
and similarly for the horizontal velocities. The radial range of
integration is taken over $r\in[R_1,R_2]$ since this is where
we find the perturbations to be confined.

\section{Linear density waves}\label{wkb}
In this paper we take a fully numerical approach using non-linear
hydrodynamic simulations. However, it will be useful to employ
existing linear density wave theory to interpret the simulations.
We list here the main definitions and results from 2D linear theory,
taken from \cite{lin11b} and \cite{shu91}.

@@

However, as noted in \cite{lin11b}, if there is an imposed
temperature gradient, as in the disc models we consider,
then $\tbg\neq 0$ in general, which corresponds to a torque
exerted by the background disc on the perturbation. We 
outline the derivation of $T_\mathrm{BG}$ in Appendix 
\ref{tbg_deriv}.

### 2014-11-07 paper/paper.tex
major updates

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we speculate it may also
  enhance low-frequency one-arm spirals in non-self-gravitating
  discs.

>>>>

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\newcommand{\varA}[1]{{\operatorname{#1}}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we speculate it may also
  affect low-frequency one-arm spirals in non-self-gravitating
  discs.

### 2014-11-07 paper/results.tex
major updates

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. The figure shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  
spiral disturbance in the inner disc \citep{lin11b}. 
%this would be a de-stabilising
% effect.  we discuss this in discussions?

>>>>

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. Note in this region, $\nu^2 - 1 +
Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
Eq. \ref{dispersion} (with the lower sign taken, i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  
spiral disturbance in the inner disc \citep{lin11b}. 
%this would be a de-stabilising
% effect.  we discuss this in discussions?

### 2014-11-07 paper/results_3d.tex
major updates

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs. We also show results from strictly
isothermal discs ($q=0$), which clearly display no growth. %why no high m
                                %modes as in 2D for iso disc?
                                %resolution?
Results from ZEUS-MP appears to be off-set from that obtained using
PLUTO. We suspect this may be due to larger numerical noise in the
ZEUS-MP code. Thus the $m=1$ spiral grows to a larger amplitude by the
end of the run. However, we measure similar growth rates,
\begin{align*}
&\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
&\gamma \simeq 0.0075\Omega_k(R_0) \quad\quad \mathrm{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations above. This may be due to the
low resolutions adopted. 
%growth rate does not depend on pert amp

@@

Visualisations of the 3D simluations are shown in Fig. \ref{3d_prelim} for
both ZEUS-MP and PLUTO. The snapshots are chosen for when the spiral
amplitudes are similar. The agreement between the two codes, as well
as with the 2D simulations, is satisfactory.

%but a longer PLUTO simulation was 
%needed to achieve similar mode amplitudes in ZEUS-MP.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.33]{figures/pdisk_020}
    }
  \end{center}
  \caption{Three-dimensional simulations using the (a) ZEUS-MP and (b)
    PLUTO. The non-axisymmetric midplane density is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height 
    from the midplane.\label{3d_prelim}}   
\end{figure}

\subsection{Vertical motion}

@@

\begin{figure}
%scale=0.41
  \includegraphics[scale=.41,clip=true,trim=0cm 1cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
  \caption{Evolution of angular momentum components in the fiducial 3D 
    simulations. The perturbation
    relative to $t=100P_0$ is shown in units of the
    initial total angular momentum $J_\mathrm{ref}$.\label{3d_angmom}} 
\end{figure}

>>>>

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs. We also show results from strictly
isothermal discs ($q=0$), which clearly display no growth. %why no high m
                                %modes as in 2D for iso disc?
                                %resolution?
Results from ZEUS-MP appears to be off-set from that obtained using
PLUTO. We suspect this may be due to larger numerical noise in the
ZEUS-MP code. Thus the $m=1$ spiral grows to a larger amplitude by the
end of the run. However, we measure similar growth rates,
\begin{align*}
&\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
&\gamma \simeq 0.0075\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations above. This may be due to the
low resolutions adopted. 
%growth rate does not depend on pert amp

@@

Visualisations of the 3D simluations are shown in
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the spiral 
amplitudes reached in the two codes are similar. Although the
non-axisymmetric density has similar amplitudes at both heights, a
coherent spiral is less evident away from the disc midplane. 
The agreement between the two codes, as well as with the 2D
simulations, is satisfactory.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.33]{figures/polarxy_dens015_z0}
      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
      0cm]{figures/polarxy_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.33]{figures/pdisk_021_z0}
      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
      0cm]{figures/pdisk_021_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The non-axisymmetric density at the midplane (left) and
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane.\label{3d_prelim}}   
\end{figure}

\subsection{Vertical motion}
In Fig. \ref{3d_vert} we measure three-dimensionality using the
parameter $\Theta$ as defined in Eq. \ref{theta}. Vertical velocities
do increase with the growth of the one-arm spiral. Larger amplitudes
are reached in ZEUS because of larger numerical noise. 
However, even 
in this case, $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
compared to horizontal motions.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
  \caption{Evolution of three-dimensionality $\Theta$, as defined by
    Eq. \ref{theta}, which corresponds to the
    average vertical kinetic energy density, scaled by 
    the average horizontal kinetic energy.\label{3d_vert}} 
\end{figure}

@@

\begin{figure}
%scale=0.41
  \includegraphics[scale=.41,clip=true,trim=0cm 1cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
  \caption{Evolution of angular momentum components in the 3D 
    simulations. The perturbation 
    relative to $t=100P_0$, during the growth of the one-arm spiral,
    is shown in units of the initial total angular momentum
    $J_\mathrm{ref}$.\label{3d_angmom}}  
\end{figure}
